---
title: 'Statistical Inference Assignment Part 2: ToothGrowth analysis'
output: html_document
---

## Loading the ToothGrowth data

```{r}
rm(list = ls())
data(ToothGrowth)
help(ToothGrowth)
```

The data was loaded with the above R code and information about the variables obtained using the 'help' function.

## Basic exploratory analyses

Information was obtained about the variables and their classes.
```{r}
str(ToothGrowth)
```

The 'dose' variable was converted into a factor variable as it will be used as a categorical variable.
```{r}
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
lapply(ToothGrowth[ , 2:3], levels)
```

The mean, sd, interquartile range, and range of the outcome 'len' was examined.
```{r}
summary(ToothGrowth$len)
sd(ToothGrowth$len)
```

The distribution of the length variable was mapped using a histogram to see shape of the distribution, and assess whether it matches the shape of a t-distribution (Figure 1).

```{r, echo = FALSE, fig.width = 6, fig.height = 4}
library(ggplot2)
g <- ggplot(data=ToothGrowth, aes(ToothGrowth$len)) + 
        geom_histogram(aes(y = ..density..), binwidth = 1, 
                       col = "#4797E5", 
                       fill = "#7EC0EE") + 
        xlab("Tooth length") + 
        ylab("Density") + 
        theme_bw() + 
        ggtitle("Histogram of tooth length") + 
        theme(plot.title = element_text(lineheight=.8, face="bold")) + 
        geom_line(aes(y = ..density.., colour = "Empirical"), stat = "density") + 
        stat_function(fun = dnorm, aes(colour = "Normal"), 
                      arg = list(mean = mean(ToothGrowth$len), sd = sd(ToothGrowth$len))) + 
        scale_colour_manual(name = "Density", values = c("black", "red"))

print(g)
```

**Figure 1.** Histogram of tooth length (the 'len' variable), mapping the density of the distribution and the normal function.

```{r, echo = FALSE}
library(xtable); library(knitr)
library(plyr)
tgSummary1 <- ddply(ToothGrowth, c("supp"), summarise,
                   n_subjects = length(len),
                   mean_len = mean(len),
                   se_len = sd(len) / sqrt(length(len))
)

levels(tgSummary1$supp) <- c("Orange juice", "Ascorbic acid")
colnames(tgSummary1) <- c("Supplement", "N", "Mean", "SE")

tgSummary2 <- ddply(ToothGrowth, c("dose"), summarise,
                    n_subjects = length(len),
                    mean_len = mean(len),
                    se_len = sd(len) / sqrt(length(len))
)

levels(tgSummary2$dose) <- c("0.5 mg", "1 mg", "2 mg")
colnames(tgSummary2) <- c("Dose", "N", "Mean", "SE")
```

By the 'supp' group, the sample size, and mean and SE for 'len' are summarised in Table 1.

```{r, echo = FALSE}
kable(tgSummary1, digits = c(3, 4))
```

**Table 1.** Summary of tooth length by supplement type.

The above was repeated for the 'dose' variable (Table 2).

```{r, echo = FALSE}
kable(tgSummary2, digits = c(3, 4))
```

**Table 2.** Summary of tooth length by dose (in mgs).

```{r, echo = FALSE}
ci1 <- transform(tgSummary1, SE_lower = Mean - SE, SE_upper = Mean + SE)
rm(tgSummary1)

ci2 <- transform(tgSummary2, SE_lower = Mean - SE, SE_upper = Mean + SE)
rm(tgSummary2)
```

Finally, a plot summarising the means and standard errors by both 'supp' and 'dose' was created (Figure 2).

```{r, echo = FALSE, fig.width = 12, fig.height = 6}
library(gridExtra)
g1 <- ggplot(data=ci1, aes(x=Supplement, y=Mean, group=Supplement, colour=Supplement)) +
        geom_bar(stat = "identity", fill="#4F87CC", colour="#4F87CC", alpha = 0.9) +
        geom_errorbar(aes(ymax=SE_upper, ymin=SE_lower), 
                      width=0.25, 
                      color = "black",
                      position=position_dodge(0.9), 
                      data=ci1) +
        ylab("Tooth length (mean)") +
        theme_bw() +
        ggtitle("Guinea Pig Tooth Growth by Supplement Type") + 
        theme(plot.title = element_text(lineheight=1.1, face="bold")) +
        theme(legend.position="none") +
        scale_y_continuous(limits=c(0, 27))

g2 <- ggplot(data=ci2, aes(x=Dose, y=Mean, group=Dose, colour=Dose)) +
    geom_bar(stat = "identity", fill="#C0D890", colour="#C0D890") +
    geom_errorbar(aes(ymax=SE_upper, ymin=SE_lower), 
                  width=0.25, 
                  color = "black",
                  position=position_dodge(0.9), 
                  data=ci2) +
    ylab("Tooth length (mean)") +
    theme_bw() +
    ggtitle("Guinea Pig Tooth Growth by Dose") + 
    theme(plot.title = element_text(lineheight=1.1, face="bold")) +
    theme(legend.position="none") +
    scale_y_continuous(limits=c(0, 27))

grid.arrange(g1, g2, nrow = 1, ncol = 2)
```

**Figure 2.** Bar graph demonstrating the mean length and standard error for each of the levels of 'supp' and 'dose'.

## Comparison of tooth length by supplement and dose
### Tooth length by supplement

The mean tooth length by supplement type (orange juice versus absorbic acid) was performed using an independent samples t-test. 

```{r}
s1 <- ToothGrowth$len[ToothGrowth$supp == "OJ"]
s2 <- ToothGrowth$len[ToothGrowth$supp == "VC"]

ttest1 <- t.test(s1, s2, alternative = "two.sided",
            paired = FALSE, 
            var.equal = FALSE)
```

The mean difference between the two groups was `r abs(mean(s1) - mean(s2))`. The 95% confidence interval around this estimate was [`r round(ttest1$conf, 3)`], which contains 0, therefore indicating that the mean tooth length between each supplement type was not significantly different. This was confirmed by the p-value, which was greater than 0.05 (`r round(ttest1$p.value, 3)`).

### Tooth length by dose
The mean tooth length by dose type was tested with two independent samples t-tests. The first, 0.5 mg versus 1 mg, is performed below:

```{r}
l1 <- ToothGrowth$len[ToothGrowth$dose == "0.5"]
l2 <- ToothGrowth$len[ToothGrowth$dose == "1"]
l3 <- ToothGrowth$len[ToothGrowth$dose == "2"]

ttest2 <- t.test(l1, l2, alternative = "two.sided",
            paired = FALSE, 
            var.equal = FALSE)
```

The mean difference between the two groups was `r abs(mean(l1) - mean(l2))`. The 95% confidence interval around this estimate was [`r round(ttest2$conf, 3)`], which does not contain 0, therefore indicating that the mean tooth length between each supplement type was significantly different at this level. This was confirmed by the p-value, which was less than 0.05 (`r format(ttest2$p.value, scientific = FALSE, digits = 1)`).

Finally, the levels 1 mg and 2 mg of dose were compared.

```{r}
ttest3 <- t.test(l2, l3, alternative = "two.sided",
            paired = FALSE, 
            var.equal = FALSE)
```

The mean difference between the two groups was `r abs(mean(l2) - mean(l3))`. The 95% confidence interval around this estimate was [`r round(ttest3$conf, 3)`], which does not contain 0, therefore indicating that the mean tooth length between each supplement type was significantly different at this level. This was confirmed by the p-value, which was less than 0.05 (`r format(ttest3$p.value, scientific = FALSE, digits = 1)`).

## Conclusions and assumptions

Overall, the data showed that the supplement type (orange juice versus ascorbic acid) did not affect the length of the guinea pig's teeth, as demonstrated by the non-significant independent samples t-test (at the 0.05 level). However, each of the dose levels of Vitamin C did. 1 mg caused a signficantly greater tooth length (`r abs(mean(l1) - mean(l2))`, 95% CI = [`r round(ttest2$conf, 3)`]) than 0.5 mg. In turn, 2 mg caused a signficantly greater tooth length (`r abs(mean(l2) - mean(l3))`, 95% CI = [`r round(ttest3$conf, 3)`]) than 1 mg.

These conclusions are only valid under the following assumptions:  
* Each of the groups are independent and data are _iid_,  
* The sample is roughly normally distriXbuted, however, the histogram (Figure 1) shows that it is likely skewed, and  
* The sample is representative of the population at hand, so that sample means and standard deviations are representative of population values.


